---
title: "ERHS 535 Final Project"
output: 
  flexdashboard::flex_dashboard:
    theme: 
      version: 4
      bootswatch: journal
    vertical_layout: fill
    orientation: columns
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(DT)
library(fuzzyjoin)
library(dplyr)
library(cowplot)
library(sf)
library(geojsonsf)
library(leaflet)

plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-01-26/plastics.csv')
```


```{r, echo = FALSE, include = FALSE}
plastics <- plastics %>%
  mutate(count_per_volunteer = (grand_total / volunteers)) %>% #normalize plastic totals by the number of volunteers
  select(!empty) %>% #don't need this column-- it's meaning is unclear
  filter(parent_company != "Grand Total") %>% #we can calculate these totals ourselves, not in the company column
  mutate(country = str_to_title(country)) #clean up country names 

#UK has two different names, fix this & fix all the other country names that conflict
plastics["country"][plastics["country"] == "United Kingdom Of Great Britain & Northern Ireland"] <- "United Kingdom"
plastics["country"][plastics["country"] == "United States Of America"] <- "United States of America"
plastics["country"][plastics["country"] == "Cote D_ivoire"] <- "Ivory Coast"
plastics["country"][plastics["country"] == "Taiwan_ Republic Of China (Roc)"] <- "Taiwan"

#what is the overall change in pollution from 2019-2020 by country 
pollution_change <- plastics %>%
  filter(!is.na(count_per_volunteer)) %>%
  group_by(country, year) %>%
  summarize(total_per_volunteer = sum(count_per_volunteer)) %>%
  pivot_wider(id_col = "country", 
              names_from = "year", 
              values_from = "total_per_volunteer") %>%
  na.omit() %>%
  mutate(difference = (`2020` - `2019`)) %>%
  mutate(color = (difference > 0))

pollution_change %>%
  pull(country) %>%
  unique() %>%
  length()
#note: we only have 41 countries where we have data for both 2019 and 2020

#what about overall change from 2019 to 2020 by plastic type
plastic_type_change <- plastics %>%
  select(!c(count_per_volunteer, parent_company, num_events)) %>%
  pivot_longer(!c(country, year, volunteers), 
               names_to = "plastic_type", values_to = "count") %>%
  filter(plastic_type != "grand_total") %>%
  mutate(count_per_volunteer = count/volunteers) %>% #gotta normalize
  mutate(year = factor(year)) %>%
  mutate(plastic_type = fct_recode(plastic_type, 
                                   "Polystyrene" = "ps",
                                   "High density polyethylene" = "hdpe",
                                   "Low density polyethylene" = "ldpe",
                                   "Polyester" = "pet", 
                                   "Polypropylene" = "pp",
                                   "PVC" = "pvc",
                                   "Other" = "o")) %>%
  filter(!is.na(count_per_volunteer)) %>%
  group_by(year, plastic_type) %>%
  summarize(total_per_volunteer = sum(count_per_volunteer)) %>%
  pivot_wider(id_col = "plastic_type", 
              names_from = "year", 
              values_from = "total_per_volunteer") %>%
  mutate(difference = (`2020` - `2019`))
```

```{r map prep, echo = FALSE, include = FALSE}
geojson_countries <- geojson_sf("https://datahub.io/core/geo-countries/r/countries.geojson")

# new df for this
plastics_map <- plastics

#replace NA values with 0
plastics_map$grand_total <- plastics_map$grand_total %>%
                        replace_na(0)

country_yearly_totals <- plastics_map %>%
  group_by(country, year) %>%
  summarise(Total = sum(grand_total))

# rename columns for merging purposes
colnames(country_yearly_totals) <- c("ADMIN", "Year", "Total")

# create grand total aggregates and volunteer count columns in seperate data frames

# to be joined and read into to the leaflet map
plastic_2019 <- country_yearly_totals %>%
  filter(Year == "2019") %>%
  select(-Year)

colnames(plastic_2019) <- c("ADMIN","Total_2019")

plastic_2020 <- country_yearly_totals %>%
  filter(Year == "2020") %>%
  select(-Year)
colnames(plastic_2020) <- c("ADMIN","Total_2020")

volunteers_2019 <- plastics_map %>%
  filter(year == "2019") %>%
  select(country,volunteers) %>%
  distinct()
colnames(volunteers_2019) <- c("ADMIN","Volunteers_2019")

volunteers_2020 <- plastics_map %>%
  filter(year == "2020") %>%
  group_by(country, volunteers) %>%
  select(country, volunteers) %>%
  distinct()
colnames(volunteers_2020) <- c("ADMIN","Volunteers_2020")

# now merge the aggregated counts with the spatial data frame
worldmap_data <- sp::merge(geojson_countries, plastic_2019, by = 'ADMIN')
worldmap_data <- sp::merge(worldmap_data, plastic_2020, by = 'ADMIN')
worldmap_data <- sp::merge(worldmap_data,  volunteers_2019, by = 'ADMIN')
worldmap_data <- sp::merge(worldmap_data,volunteers_2020, by = 'ADMIN')

mytext <- paste(
  "Country: ", worldmap_data$ADMIN, 
  "<br/> 2020 - Total Plastic: ", worldmap_data$Total_2020, "&emsp;Total Volunteers: ", worldmap_data$Volunteers_2020,
  "<br/> 2019 - Total Plastic: ", worldmap_data$Total_2019, "&emsp;Total Volunteers: ", worldmap_data$Volunteers_2019,
  sep = " "
) %>%
  lapply(htmltools::HTML)

pal <- colorNumeric(palette= "YlOrBr",
                    domain = log(worldmap_data$Total_2020), n=5)
```

The data for this project come from [#breakfreefromplastic](https://www.breakfreefromplastic.org/). This group organizes citizen science plastic cleanup events around the world. Although their overall goal is to hold companies accountable for plastic waste, their datasets hold a wealth of additional information. In this dashboard, we investigate the distribution of plastic pollution by parent company, country, year and plastic type. Importantly, we calculated an "index of plastic pollution" that takes into account the amount of effort that went into a given cleanup event, as the number of participants affected the amount plastic that was collected. This allows us to compare pollution across all of the variables in the dataset.

$\text{count per volunteer} = \frac{\text{plastic count}_{}}{\text{number of volunteers}_{}}$

Column {.tabset .tabset-fade}
-------------------------------------

### World maps of plastic pollution in 2019 and 2020
```{r}
leaflet(worldmap_data) %>% 
  addTiles()  %>% 
  setView( lat=10, lng=0 , zoom=2) %>%
  addPolygons( 
    fillColor = ~pal(log(worldmap_data$Total_2020)), 
    stroke=TRUE, 
    fillOpacity = 0.9, 
    color="white", 
    weight=0.3,
    label = mytext,
    labelOptions = labelOptions( 
      style = list("font-weight" = "normal", padding = "3px 8px"), 
      textsize = "13px", 
      direction = "auto"
    )
  ) %>%
  addLegend( pal= pal, values= ~log(worldmap_data$Total_2020), opacity=0.9, title = "Total plastic pollution", position = "bottomleft" )
```

### Plastic pollution by parent company

```{r}
#Trying to fuzzy join
x <- plastics %>%
  select(year, parent_company)
y <- plastics

fuzzy_plastic <- plastics %>%
  select(year, parent_company) %>%
  fuzzy_full_join(x, y, by = c("parent_company" = "parent_company"), match_fun = list(`==`))


#Original code for parent company table
plastic_companies <- plastics %>%
  select(year, parent_company, grand_total) %>%
  filter(parent_company != "Grand Total") %>%
  #filter(year == "2019") %>%
  filter(parent_company != "null") %>%
  filter(parent_company != "NULL") %>%
  filter(parent_company != "Unbranded") %>%
  filter(parent_company != "#ERROR!") %>%
  na.omit() %>%
  group_by(year, parent_company) %>%
  summarise(ggrand = sum(grand_total)) %>%
  arrange(desc(ggrand)) %>%
  ungroup() %>%
  rename(Company = parent_company) %>%
  rename(Total = ggrand) %>%
  pivot_wider(id_cols = "Company",
              names_from = "year",
              values_from = "Total") %>%
  datatable(options = list(pagelength = 12), caption = "Total Collected Pieces of Plastic Pollution by Parent Company in 2019 and 2020")
plastic_companies
```


### Change in plastic pollution from 2019 to 2020

```{r}
p <- pollution_change %>%
  ggplot() +
  geom_point(aes(x = difference, 
                 y = reorder(country, difference),
                 color = color)) +
  # scale_color_brewer(palette = "Pastel1") +
  scale_color_manual(values = c("#4DBBD599", "#E64B3599")) +
  geom_segment(aes(y = country, yend = country,
                   x = 0, xend = difference)) +
  labs(y = "", 
       x = "Change in pollution from 2019 to 2020 \n(pieces of plastic per volunteer)") +
  theme_classic() +
  theme(legend.position = "none")

p2 <- plastic_type_change %>%
  ggplot() +
  geom_bar(aes(x = difference, 
                 y = reorder(plastic_type, difference)), 
           fill = "#E64B3599", stat = "identity") +
  labs(y = "", x = "Change in pollution \nby plastic type") +
  theme_bw() +
  theme(legend.position = "none")

ggdraw(p) +
  draw_plot(p2, .5, .2, .5, .5)
```

